---
title: "Stats for llm-Players"
output:
  pdf_document: default
  html_notebook: default
editor_options: 
  chunk_output_type: inline
---
# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(echo = F)
```

```{r}
rm(list = ls())
library(ggplot2)
library(dplyr)
library(viridisLite)
library(tidyr)
library(scales)
getwd()

source("scripts/load_data.R")

colors.opinion <- c("red", "orange", "grey", "cyan", "blue")
colors.actor <- c("turquoise3", "thistle4")
colors.quality <- c("#458B00", "darkgoldenrod3", "#8B1C62", "red4")

interpretation.opponent_opinion_human_or_ai <- as.factor(c(
  "Definitely human",
  "Probably human",
  "Don't know",
  "Probably ai",
  "Definitely ai"
))
```

```{r}
  
number_of_test_games <- tests[!tests$is_equivalence_test,] |>
  group_by(
    llm_or_human,
  ) |>
  summarise(
    n = n_distinct(game_id)
  )
number_of_mutant_games <- mutants |>
  group_by(
    llm_or_human,
  ) |>
  summarise(
    n = n_distinct(game_id)
  )

deftests <- tests |> 
  filter(!is_equivalence_test, !fails_against_cut)
```

# Killrates

A "Killrate" is calculated for every single test. It describes the percentage of 
mutants within a specific population that would be killed by this test among 
those that are covered by this test.

"existing" and "future" mutants refers to those mutants within the same game
as the test that were submitted before/after the test was submitted.

```{r}

deftests |>
  pivot_longer(
    cols = c(kill_rate, llm_kill_rate, human_kill_rate, existing_kill_rate, future_kill_rate),
    names_to = "kill_rate_type",
    values_to = "kill_rate_value"
  ) |>
  mutate(kill_rate_type = factor(kill_rate_type, 
                                 levels = c(
                                   "kill_rate", 
                                   "human_kill_rate", 
                                   "llm_kill_rate", 
                                   "existing_kill_rate",
                                   "future_kill_rate"),
                                 labels = c(
                                   "Total kill rate",
                                   "Against human mutants",
                                   "Against llm mutants",
                                   "Against existing mutants",
                                   "Against future mutants"
                                 ))) |>
  ggplot(aes(x = llm_or_human, y = kill_rate_value, fill = opponent_llm_or_human)) + 
    geom_boxplot(position = "dodge") + 
    xlab("Defender") +
    scale_y_continuous(name = "Kill rate", labels = percent) + 
    scale_fill_manual(name = "Attacker", values = colors.actor) +
    facet_wrap( ~ kill_rate_type)
```

## Finding new mutants
How successful are tests in finding mutants that have previously been undetected?
This counts the absolute number of mutants, so there are no percentages.

Most tests only directly kill 0 or 1 mutants.

```{r}
deftests |>
  ggplot(aes(x = llm_or_human, y = fresh_killed_mutants, fill = opponent_llm_or_human)) +
  geom_boxplot() +
  xlab("Defender") + 
  ylab("Number of killed mutants") +
  scale_fill_manual(values = colors.actor, name = "Attacker")

```

Comparing averages shows clearer results. For both classes, human-written
tests kill more hitherto surviving mutants. Tests also have kill more new
LLM-written mutants.

Possible explanations for this: Human defenders write fewer tests than LLM 
defenders, so every test has more mutants available to kill.

```{r}
deftests |>
  summarise(fresh_killed_mutants = mean(fresh_killed_mutants), .by = c(llm_or_human, opponent_llm_or_human, cut)) |>
  ggplot(aes(x = llm_or_human, y = fresh_killed_mutants, fill = opponent_llm_or_human)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(fresh_killed_mutants, digits = 2)), position = position_dodge(0.9), vjust = 1.5) +
  scale_fill_manual(values = colors.actor, name = "Attacker") +
  ylab("Number of fresh mutant kills") +
  xlab("Defender") +
  facet_wrap( ~ cut)
```



# Evasion rate

An evasion rate is the equivalent of a kill rate for mutants: How many of the
tests in a specific population that cover this test could this mutant survive?

```{r}
mutants |>
  pivot_longer(
    cols = c(evasion_rate, llm_evasion_rate, human_evasion_rate, existing_evasion_rate, future_evasion_rate),
    names_to = "evasion_rate_type",
    values_to = "evasion_rate_value"
  ) |>
  mutate(evasion_rate_type = factor(evasion_rate_type, 
                                 levels = c(
                                   "evasion_rate", 
                                   "human_evasion_rate", 
                                   "llm_evasion_rate", 
                                   "existing_evasion_rate",
                                   "future_evasion_rate"),
                                 labels = c(
                                   "Total evasion rate",
                                   "Against human tests",
                                   "Against llm tests",
                                   "Against existing tests",
                                   "Against future tests"
                                 ))) |>
  ggplot(aes(x = llm_or_human, y = evasion_rate_value, fill = opponent_llm_or_human)) + 
    geom_boxplot(position = "dodge") + 
    xlab("Attacker") +
    scale_y_continuous(name = "Evasion rate", labels = percent) + 
    scale_fill_manual(name = "Defender", values = colors.actor) +
    facet_wrap( ~ evasion_rate_type)
```


# Points

## Points of individual tests

The following plots show the distribution of points for tests and mutants.
As we can see, most tests score very low, with some exceptionally high outliers.

```{r}
point_boxplot <- function(.data, author = "Author", opponent = "Opponent") {
  .data |>
    ggplot(aes(x = llm_or_human, y = points, fill = opponent_llm_or_human)) + 
    geom_boxplot() +
    scale_fill_manual(values = colors.actor, name = opponent) +
    xlab(author) +
    ylab("Points")
}

deftests |>
  point_boxplot()
```

This plot shows the percentage of tests with at least one point. When playing
against an LLM attacker, humans and LLM defenders get at least one point
at almost the exact same rate, but against humans, human defenders score more
often.

```{r}
deftests |>
  mutate(has_points = points > 0) |>
  group_by(llm_or_human, opponent_llm_or_human) |>
  summarise(has_points = mean(has_points)) |>
  ggplot(aes(y = has_points, x = llm_or_human, fill = opponent_llm_or_human)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = percent(after_stat(y))), vjust = 1.5, position = position_dodge(0.9)) +
  scale_fill_manual(values = colors.actor, name = "Attacker") +
  scale_y_continuous(labels = percent, name = NULL)+ 
  xlab("Defender") +
  ggtitle("Tests with at least one point")
```
The average human-written tests scores higher than the average LLM-written test.

```{r}
deftests |>
  group_by(llm_or_human, opponent_llm_or_human) |>
  summarise(mean_points = mean(points)) |>
  ggplot(aes(x = llm_or_human, y = mean_points, fill = opponent_llm_or_human)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = mean_points |> round(2)), 
    position = position_dodge(0.9),
    vjust = 1.5
    ) +
  xlab("Defender") +
  ylab("Points") +
  ggtitle("Average test points") + 
  scale_fill_manual(name = "Attacker", values = colors.actor)
```
## Points of individual mutants

Same as with tests, mutant points are usually very low, with some outliers.

```{r}
mutants |>
  point_boxplot()
```
Humans are somewhat more successful in writing mutants that garner at least one
point. Getting at least a single point against a human defender seems to be much
easier for both LLM and human attackers, compared with an LLM defender.

```{r}
mutants |>
  mutate(has_points = points > 0) |>
  group_by(llm_or_human, opponent_llm_or_human) |>
  summarise(has_points = mean(has_points)) |>
  ggplot(aes(y = has_points, x = llm_or_human, fill = opponent_llm_or_human)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = percent(after_stat(y))), vjust = 1.5, position = position_dodge(0.9)) +
  scale_fill_manual(values = colors.actor, name = "Defender") +
  scale_y_continuous(labels = percent, name = NULL)+ 
  xlab("Attacker") +
  ggtitle("Mutants with at least one point")
```
In contrast, the average number of points is higher when playing against an
LLM defender compared to a human defender. Average mutant points are
generally much higher for human attackers.

```{r}
deftests |>
  group_by(llm_or_human, opponent_llm_or_human) |>
  summarise(mean_points = mean(points)) |>
  ggplot(aes(x = llm_or_human, y = mean_points, fill = opponent_llm_or_human)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = mean_points |> round(2)), 
    position = position_dodge(0.9),
    vjust = 1.5
    ) +
  xlab("Attacker") +
  ylab("Points") +
  ggtitle("Average mutant points") + 
  scale_fill_manual(name = "Defender", values = colors.actor)
```


# Relationship between points and killrates

While there is no clear pattern for the relationship between the points
earned in-game by test and their total "kill-rate" against all mutants submitted
in the study, mutants that earn many points tend to have higher "evasion-rates"
against all tests submitted in the study.

Mutants that were declared as equivalent while not being killed by any test
were excluded from the sample.

```{r}
deftests |> 
  mutate(
    point_ntil = ntile(points, n = 10)
  ) |>
  group_by(point_ntil) |>
  ggplot(aes(y = kill_rate, x = factor(point_ntil), fill = llm_or_human)) +
    geom_boxplot() +
    scale_fill_manual(values = colors.actor, name = "Defender") + 
    ggtitle("Kill rates vs Points for tests") +
    xlab("Point deciles") + 
    ylab("Total kill rate") +
    scale_y_continuous(labels = percent)

```

```{r}
mutants |> 
  filter(!(!has_been_killed &declared_equivalent)) |>
  mutate(
    point_ntil = ntile(points, n = 10)
  ) |>
  group_by(point_ntil) |>
  ggplot(aes(y = evasion_rate, x = factor(point_ntil), fill = llm_or_human)) +
    geom_boxplot() +
    scale_fill_manual(values = colors.actor, name = "Attacker") + 
    ggtitle("Evasion rates vs Points for mutants") +
    xlab("Point deciles") + 
    ylab("Total evasion rate") +
    scale_y_continuous(labels = percent)
```

```{r}
ggplot(mutants, aes(x = points, y = evasion_rate)) +
  geom_point() + 
  geom_smooth(formula = y ~x, method = loess) +
  scale_y_continuous(limits = c(0,1), labels = percent, name = "Global evasion rate") + 
  xlab("Points")
```

# Submission timelines

These plots show the average number of tests/mutants submitted per minute per game.

We can see that LLM defenders submitted significantly more tests per minute than
human defenders, showing that the cooldown period was configured to be too low.

Submission rates for LLM defenders go down with time. It seems that the initial
"test suite" tests require re-generation less often than the uniquely written tests.

```{r}
timelineplot <- function(.data) {
  .data |>
  mutate(seconds_since_gamestart = floor(seconds_since_gamestart / 60)) |>
    rename(`Minutes since gamestart` = seconds_since_gamestart) |>
    mutate(
      part = as.numeric(ifelse(llm_or_human == "LLM",
                    1 / number_of_test_games[number_of_test_games$llm_or_human == "LLM","n"],
                    1 / number_of_test_games[number_of_test_games$llm_or_human == "Human","n"]
                    ))
    ) |>
    group_by(`Minutes since gamestart`, llm_or_human) |>
    summarise(
      part = sum(part)
    ) |>
  ggplot(aes(x = `Minutes since gamestart`, y = part, colour = llm_or_human)) +
    geom_line()
}

deftests |>
  timelineplot() +
  ylab("Test submissions per game and minute") +
  scale_colour_manual(values = colors.actor, name = "Defender")
```

Submission rates for mutants go down over time for both LLM and Human Attackers.
Since LLM attacker attempt to submit a new mutant every 30 seconds, this must
be related to higher rejection rates. A possible reason for that is the requirement
of mutants to be unique, there will be more collisions later in the game.

```{r}
mutants |>
  timelineplot() +
  ylab("Mutant submissions per game and minute") +
  scale_colour_manual(values = colors.actor, name = "Attacker")
```

# LLM messages until success

LLM players can try to fix their generated tests/mutants if they failed validation.
These plots show the number of successfully submitted tests/mutants per minute,
grouped by the number of messages to the LLM that were necessary for this submission.

Some tests have no directly associated number of messages. Those tests were generated
as part of the initial test suite and were successfully submitted on first try.


```{r}
deftests|>
  filter(llm_or_human == "LLM") |>
  mutate(number_of_messages = ifelse(is.na(number_of_messages), 0, number_of_messages)) |>
  mutate(number_of_messages = factor(number_of_messages,
                                     levels = c(0, 1,2,3,4),
                                     labels = c("Suite", "1", "2", "3", "4"))) |>
  mutate(seconds_since_gamestart = seconds_since_gamestart / 60) |>
  rename(`Minutes since gamestart` = seconds_since_gamestart) |>
  group_by(number_of_messages) |>
  ggplot(aes(x = `Minutes since gamestart`, colour = number_of_messages)) +
    geom_freqpoly(binwidth = 1) + 
    ylab("Tests per minute") +
    scale_colour_manual(name = "Number of messages", values = c("black", colors.quality))
  
```

```{r}
mutants|>
  filter(llm_or_human == "LLM") |>
  mutate(number_of_messages = factor(number_of_messages,
                                     levels = c(0, 1,2,3,4),
                                     labels = c("Suite", "1", "2", "3", "4"))) |>
  mutate(seconds_since_gamestart = seconds_since_gamestart / 60) |>
  rename(`Minutes since gamestart` = seconds_since_gamestart) |>
  group_by(number_of_messages) |>
  ggplot(aes(x = `Minutes since gamestart`, colour = number_of_messages)) +
    geom_freqpoly(binwidth = 1) + 
    ylab("Mutants per minute") +
    scale_colour_manual(name = "Number of messages", values = c(colors.quality))
```



# Turing tests

LLM-generated tests are almost never confused for human-written tests,
but mutants are hardly distinguishable.

```{r}
deftests |>
  filter(opponent_llm_or_human == "Human") |>
  group_by(game_id, username, opponent_opinion_human_or_ai, llm_or_human) |>
  mutate(opponent_opinion_human_or_ai = factor(opponent_opinion_human_or_ai, levels = 1:5, labels = interpretation.opponent_opinion_human_or_ai)) |>
  summarise(n = n()) |>
  ggplot(aes(x = llm_or_human, y = n, fill = opponent_opinion_human_or_ai)) + 
    geom_col(position = "fill", name = "Proportion") +
    scale_fill_manual(name = "Attacker's guess", values = colors.opinion) +
    xlab("Defender") +
    ylab("Proportion") +
    ggtitle("Defender Turing test") + 
    scale_y_continuous(labels = percent)

```

```{r}
mutants |>
  filter(opponent_llm_or_human == "Human") |>
  group_by(game_id, username, opponent_opinion_human_or_ai, llm_or_human) |>
  mutate(opponent_opinion_human_or_ai = factor(opponent_opinion_human_or_ai, levels = 1:5, labels = interpretation.opponent_opinion_human_or_ai)) |>
  summarise(n = n()) |>
  ggplot(aes(x = llm_or_human, y = n, fill = opponent_opinion_human_or_ai)) + 
    geom_col(position = "fill", name = "Proportion") +
    scale_fill_manual(name = "Defender's guess", values = colors.opinion) +
    scale_y_continuous(labels = percent) +
    xlab("Attacker") +
    ylab("Proportion") + 
    ggtitle("Attacker Turing test")
```
# Stillborn mutants

A stillborn mutant is a mutant that was killed by a test that was already in the game.

## Proportions
This plot shows the proportions of stillborn mutants within a population.
We can see that LLM-generated mutants are more likely to be stillborn than
human-written mutants, and that mutants are much more likely to be killed
on submission when playing against an LLM player compared to a human player.

This last point can possibly be explained by the fact that LLM defenders first 
generate a complete test suite before writing individual tests.

```{r}
mutants |>
  group_by(llm_or_human, opponent_llm_or_human) |>
  summarise(
    stillborn = mean(is_stillborn)
  ) |>
  ggplot(aes(x = llm_or_human, y = stillborn, fill = opponent_llm_or_human)) +
    geom_col(position = "dodge") +
    geom_text(
      aes(label = percent(stillborn)), 
      position = position_dodge(0.9),
      vjust = 1.5,
    ) +
    scale_fill_manual(name = "Defender", values = colors.actor) + 
    xlab("Attacker") + 
    ylab("Percentage") + 
    ggtitle("Percentages of stillborn mutants") +
    scale_y_continuous(labels = percent)

```
There are, however, only very slight differences in total evasion rates
between stillborn and not-stillborn mutants.

```{r}
ggplot(data = mutants, aes(x = is_stillborn, y = evasion_rate, fill = llm_or_human)) +
  geom_boxplot(position = "dodge") +
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  scale_y_continuous(labels = percent, name = "Evasion rate") +
  scale_x_discrete(labels = c("Not stillborn", "Stillborn"),) +
  xlab(NULL) +
  ggtitle("Total evasion rates of stillborn and not-stillborn mutants")
```



# Equivalences
The following plots show the percentage of mutants that could not be killed by any test
in the entire study among their respective population.

For the purpose of this analysis, we assume that all mutants that could not be killed
by any test in the study are actually equivalent. 

We can see that at least 20% of human-written mutants that were declared as equivalent
after being claimed as equivalent by another human were actually killable. 

In contrast, at least 57% of LLM-written tests that 
were accepted as equivalent after being claimed as equivalent by a human were
killable. 

In LLM-vs-LLM-games, almost every equivalent test was marked as equivalent.
This is because in LLM-vs-LLM games, after the normal time ends, players
stop the normal test/mutant submission, the defender claims every surviving
mutant as equivalent, and the attacker is given the opportunity to defend
against that. This procedure was chosen because without it, many equivalent
tests with just shy of 10 points would in total provide a massive point 
advantage to the attacker. LLM-vs-LLM games are therefore not comparable to the
other combinations.
```{r}

mutants |>
  group_by(
    llm_or_human,
    declared_equivalent,
    opponent_llm_or_human
  ) |>
  summarise(
    not_killed = mean(!has_been_killed)
  ) |>
  mutate(
    opponent_llm_or_human = paste(opponent_llm_or_human, "Defender")
  ) |>
  ggplot(
    aes(x = declared_equivalent, y = not_killed, fill = llm_or_human)
    ) +
    geom_col(position = "dodge") +
    geom_text(aes(
      label = percent(not_killed),
      vjust = -0.4
      ), position = position_dodge(0.9)) +
    facet_wrap( ~ opponent_llm_or_human,) + 
    scale_y_continuous(labels = percent, name = "Equivalent mutants") + 
    scale_x_discrete(labels = c("Not declared\n equivalent", "Declared equivalent")) +
    scale_fill_manual(name = "Attacker", values = colors.actor) +
    xlab(NULL)

  
```
```{r}
mutants |>
  group_by(
    llm_or_human,
    opponent_llm_or_human,
    has_been_killed
  ) |>
  summarise(
    declared_equivalent = mean(declared_equivalent)
  ) |>
  mutate(
    opponent_llm_or_human = paste(opponent_llm_or_human, "Defender")
  ) |>
  ggplot(
    aes(x = has_been_killed, y = declared_equivalent, fill = llm_or_human)
    ) +
    geom_col(position = "dodge") +
    geom_text(aes(
      label = percent(declared_equivalent),
      vjust = -0.4
      ), position = position_dodge(0.9)) +
    facet_wrap( ~ opponent_llm_or_human,) + 
    scale_y_continuous(labels = percent, name = "Mutants declared as equivalent") + 
    scale_x_discrete(labels = c("Equivalent", "Not equivalent")) +
    scale_fill_manual(name = "Attacker", values = colors.actor) +
    ggtitle("Equivalent mutants") + 
    xlab(NULL)
```
# Survival rates
The following plots show the percentage of mutants within the specified 
population that survive until the end of their game.

Human-written mutants have much lower survival rates when the defender is
an LLM player compared to a human player.

The survival rate is 0% for LLM-vs-LLM games, because all surviving mutants are
automatically claimed as equivalent at the end of the game.

```{r}
mutants |>
  group_by(llm_or_human, opponent_llm_or_human) |>
  summarise(`Survival rate` = mean(survived)) |>
  ggplot(aes(y = `Survival rate`, x = llm_or_human, fill = opponent_llm_or_human)) +
  scale_fill_manual(values = colors.actor, name = "Defender") +
  xlab("Attacker") + 
  scale_y_continuous(name = "Survival rate", labels = percent) +
  geom_col(position = "dodge") + 
  geom_text(aes(label = percent(`Survival rate`)), vjust = -0.2, position = position_dodge(0.9)) + 
  ggtitle("Mutant survival rates")
```
```{r}
mutants |>
  filter(!is_stillborn) |>
  group_by(llm_or_human, opponent_llm_or_human) |>
  summarise(`Survival rate` = mean(survived)) |>
  ggplot(aes(y = `Survival rate`, x = llm_or_human, fill = opponent_llm_or_human)) +
  scale_fill_manual(values = colors.actor, name = "Defender") +
  xlab("Attacker") + 
  scale_y_continuous(name = "Survival rate", labels = percent) +
  geom_col(position = "dodge") + 
  geom_text(aes(label = percent(`Survival rate`)), vjust = -0.2, position = position_dodge(0.9)) +
  ggtitle("Mutant survival rates (excluding stillborn mutants)")
```

# Line coverage

This plot shows the number of lines being covered by a test.
LLM generated tests cover slightly more lines than human written tests.

```{r}
deftests |>
  ggplot(aes(x = llm_or_human, y = number_lines_covered, fill = opponent_llm_or_human)) +
  geom_boxplot() +
  scale_fill_manual(values = colors.actor, name = "Attacker") +
  facet_wrap( ~ cut) +
  xlab("Defender") + 
  ylab("Covered lines") + 
  ggtitle("Test coverage")
```
For mutants, the difference is more pronounced. Most human mutants only change a
single line, while LLM mutants often change multiple lines.

However, this counts all changes, including changes to comments which were
illegal for human players, but permitted to LLM players.

```{r}
mutants |>
  ggplot(aes(x = llm_or_human, y = number_mutated_lines, fill = opponent_llm_or_human)) +
  geom_boxplot() +
  scale_fill_manual(values = colors.actor, name = "Defender") +
  xlab("Attacker") + 
  ylab("Mutated lines") + 
  ggtitle("Mutated lines") + 
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = after_stat(y)),
    vjust = -0.5,
    position = position_dodge(1)
  ) +
  facet_wrap( ~ cut)
```
## Which lines were covered and mutated?

TODO: Normalise for per-game averages, compare mutations and coverage
in the same plot
```{r}

coverage_plot <- function(.data, lines_variable_name) {
  .data |>
    rename(.lines = {{lines_variable_name}}) |>
    select(llm_or_human, opponent_llm_or_human, .lines) |>
    separate_longer_delim(.lines, delim=",") |>
    group_by(llm_or_human, opponent_llm_or_human, .lines) |>
    summarise(n = n() ) |>
    ungroup() |>
    mutate(.lines = as.numeric(.lines)) |>
    ggplot(aes(x = .lines, y = n)) +
    geom_col() +
    facet_grid(llm_or_human ~ opponent_llm_or_human) +
    scale_y_continuous()
}

deftests |>
  filter(cut == "CharRange") |>
  coverage_plot(lines_covered) +
  ggtitle("Coverage for CharRange")
```

```{r}
deftests |>
  filter(cut == "ByteVector") |>
  coverage_plot(lines_covered) +
  ggtitle("Coverage for ByteVector")
```

```{r}
mutants |>
  filter(cut == "CharRange") |>
  coverage_plot(mutatedlines) +
  ggtitle("Mutations for CharRange")
```

```{r}
mutants |>
  filter(cut == "ByteVector") |>
  coverage_plot(mutatedlines) +
  ggtitle("Mutations for CharRange")
```




# Game statistics
The following statistics are about entire games, not about individual tests
and mutants.

```{r}
defender_games <- deftests |>
  group_by(
    game_id, 
    username, 
    across(starts_with("author_")), 
    across(starts_with("opponent_")),
    llm_or_human,
    cut,
    round
    ) |>
  summarise(
    points = sum(points)
  )

attacker_games <- mutants |>
  group_by(
    game_id, 
    username, 
    across(starts_with("author_")), 
    across(starts_with("opponent_")),
    llm_or_human,
    cut,
    round
    ) |>
  summarise(
    points = sum(points)
  )
```
## Total points

These plots show the total number of points players earned in a game.
There is no strong pattern except that defenders tend to earn more points
against LLM attackers than against human attackers, while attackers tend
to earn more points against human defenders than against LLM defenders.

```{r}
defender_games |>
  ggplot(aes(y = points, x = llm_or_human, fill = opponent_llm_or_human)) +
  geom_boxplot() +
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  ylab("Points") +
  xlab("Defender") +
  ggtitle("Defender games") + 
  facet_wrap( ~ cut)
```

```{r}
attacker_games |>
  ggplot(aes(y = points, x = llm_or_human, fill = opponent_llm_or_human)) +
  geom_boxplot(position = position_dodge(0.9)) +
  scale_fill_manual(name = "Defender", values = colors.actor) +
  ylab("Points") +
  xlab("Attacker") +
  ggtitle("Attacker games") +
  facet_wrap( ~ cut)
```

## Winrates

Generally, attackers have a very high win rate, at about 80%. The only exception
is in LLM-vs-LLM games, where it falls down to just 22%. 

This difference could be the effect of a unique mechanic of LLM-vs-LLM games,
where all surviving mutants are claimed as equivalent. As seen above, LLM
attackers are much more likely to fail to prove that a mutant is not equivalent,
resulting in many more lost points due to lost equivalence duels.

(Defender winrates and attacker winrates are essentially the same data,
reversed. There is not a single draw in the data set.) 

```{r}
games <- attacker_games |>
  inner_join(defender_games, join_by(game_id, cut, round), suffix = c(".attacker", ".defender"))|>
  select(!starts_with("opponent"))

games |>
  group_by(llm_or_human.attacker, llm_or_human.defender) |>
  summarise(
    defender_winrate = mean(points.defender > points.attacker)
    ) |>
  ggplot(aes(x = llm_or_human.defender, fill = llm_or_human.attacker, 
             y = defender_winrate, label = percent(defender_winrate))) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = colors.actor, name = "Attacker") + 
  xlab("Defender") +
  ggtitle("Defender winrates") +
  scale_y_continuous(labels = percent, name = "Defender winrate") +
  geom_text(position = position_dodge(0.9), vjust = 1.5)
```

```{r}
games |>
  group_by(llm_or_human.attacker, llm_or_human.defender) |>
  summarise(
    attacker_winrate = mean(points.defender < points.attacker)
    ) |>
  ggplot(aes(x = llm_or_human.attacker, fill = llm_or_human.defender, 
             y = attacker_winrate, label = percent(attacker_winrate))) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = colors.actor, name = "Defender") + 
  xlab("Attacker") +
  ggtitle("Attacker winrates") +
  scale_y_continuous(labels = percent, name = "Attacker winrate") +
  geom_text(position = position_dodge(0.9), vjust = 1.5)
```

## Point ratios

Win/Lose is a binary value that doesn't show the degree to which a player has
won or lost. For this, we will look at the ratio of defender points vs attacker
points.

```{r}
games |>
  group_by(llm_or_human.attacker, llm_or_human.defender) |>
  mutate(point_ratio = points.defender / ifelse(points.attacker != 0, points.attacker, 1)) |>
  ggplot(aes(x = llm_or_human.defender, fill = llm_or_human.attacker, y = point_ratio)) +
  geom_boxplot() +
  scale_fill_manual(values = colors.actor, name = "Attacker") +
  xlab("Defender") +
  ylab("Defender points / attacker points")
```



