---
title: "Stats for llm-Players"
output:
  html_notebook: default
  pdf_document: default
---
# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r}
library(ggplot2)
library(dplyr)
library(viridisLite)
library(tidyr)
library(scales)
getwd()
detach.dfs <- function() {
  for (p in search()) {
    if (grepl("tests|mutants", p)) {
      detach(p, character.only = TRUE)
    }
  }
}

colors.opinion <- c("red", "orange", "grey", "cyan", "blue")
colors.actor <- c("turquoise3", "thistle3")
colors.quality <- c("#458B00", "darkgoldenrod3", "#8B1C62", "red4")

interpretation.opponent_opinion_human_or_ai <- as.factor(c(
  "Definitely human",
  "Probably human",
  "Don't know",
  "Probably ai",
  "Definitely ai"
))

levels.degree <- c(
  "Bachelor Informatik", 
  "Bachelor Internet Computing", 
  "Lehramt Informatik",
  "Artificial Intelligence")
levels.cut <- c(
  "ByteVector",
  "CharRange"
)

common_variables <- function(.data) {
  mutate(
    .data,
    llm_or_human = factor(llm_or_human, levels = c("LLM", "Human")),
    opponent_llm_or_human = factor(opponent_llm_or_human, levels = c("LLM", "Human")),
    cut = factor(cut, levels = levels.cut),
    author_degree = factor(author_degree, levels = levels.degree),
    opponent_degree = factor(opponent_degree, levels = levels.degree),
    experiment_round = factor(experiment_round),
    author_opinion_judged_programming_skill = factor(author_opinion_judged_programming_skill),
    author_opinion_judged_original = factor(author_opinion_judged_original),
    author_opinion_challenged = factor(author_opinion_challenged),
    author_opinion_fun = factor(author_opinion_fun),
    author_opinion_human_or_ai = factor(author_opinion_human_or_ai),
    opponent_opinion_judged_programming_skill = factor(opponent_opinion_judged_programming_skill),
    opponent_opinion_judged_original = factor(opponent_opinion_judged_original),
    opponent_opinion_challenged = factor(opponent_opinion_challenged),
    opponent_opinion_fun = factor(opponent_opinion_fun),
    opponent_opinion_human_or_ai = factor(opponent_opinion_human_or_ai),
  )
}
```

```{r}
tests <- read.csv("test_data.csv") |>
  common_variables() |>
  mutate(
    is_equivalence_test = is_equivalence_test == "YES",
    kill_rate = (human_mutants_killed + llm_mutants_killed) / (human_mutants_killed + human_mutants_ignored + llm_mutants_killed + llm_mutants_ignored) |>
      replace_na(0),
    human_kill_rate = (human_mutants_killed / (human_mutants_killed + human_mutants_ignored)) |>
      replace_na(0),
    llm_kill_rate = (llm_mutants_killed / (llm_mutants_killed + llm_mutants_ignored)) |>
      replace_na(0),
    existing_kill_rate = (existing_mutants_killed / (existing_mutants_killed + existing_mutants_ignored)) |>
      replace_na(0),
    future_kill_rate = (future_mutants_killed / (future_mutants_killed + future_mutants_ignored)) |>
      replace_na(0),
  )

mutants <- read.csv("mutant_data.csv") |>
  common_variables() |>
  mutate(
    stillborn = stillborn == "YES",
    survived = survived == "YES",
    declared_equivalent = declared_equivalent == "YES",
    has_been_killed = human_tests_killed_by + llm_tests_killed_by > 0,
    
    evasion_rate = (human_tests_survived + llm_tests_survived) / (human_tests_survived + human_tests_killed_by + llm_tests_survived + llm_tests_killed_by) |>
      replace_na(0),
    human_evasion_rate = (human_tests_survived / (human_tests_survived + human_tests_killed_by)) |>
      replace_na(0),
    llm_evasion_rate = (llm_tests_survived / (llm_tests_survived + llm_tests_killed_by)) |>
      replace_na(0),
    existing_evasion_rate = (existing_tests_survived / (existing_tests_survived + existing_tests_killed_by)) |>
      replace_na(0),
    future_evasion_rate = (future_tests_survived / (future_tests_survived + future_tests_killed_by)) |>
      replace_na(0),
  )
  
number_of_test_games <- tests[!tests$is_equivalence_test,] |>
  group_by(
    llm_or_human,
  ) |>
  summarise(
    n = n_distinct(game_id)
  )
number_of_mutant_games <- mutants |>
  group_by(
    llm_or_human,
  ) |>
  summarise(
    n = n_distinct(game_id)
  )
```

# Killrates

A "Killrate" is calculated for every single test. It describes the percentage of 
mutants within a specific population that would be killed by this test among 
those that are covered by this test.

"existing" and "future" mutants refers to those mutants within the same game
as the test that were submitted before/after the test was submitted.

```{r}

tmp <- tests |>
  pivot_longer(
    cols = c(kill_rate, llm_kill_rate, human_kill_rate, existing_kill_rate, future_kill_rate),
    names_to = "kill_rate_type",
    values_to = "kill_rate_value"
  ) |>
  mutate(kill_rate_type = factor(kill_rate_type, 
                                 levels = c(
                                   "kill_rate", 
                                   "human_kill_rate", 
                                   "llm_kill_rate", 
                                   "existing_kill_rate",
                                   "future_kill_rate"),
                                 labels = c(
                                   "Total kill rate",
                                   "Against human mutants",
                                   "Against llm mutants",
                                   "Against existing mutants",
                                   "Against future mutants"
                                 )))

ggplot(data = tmp, 
       aes(x = llm_or_human, y = kill_rate_value, fill = opponent_llm_or_human)
       ) + 
  geom_boxplot(position = "dodge") + 
  xlab("Defender") +
  scale_y_continuous(name = "Kill rate", labels = percent) + 
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  facet_wrap( ~ kill_rate_type)
```

```{r}
plot(tests$points ~ tests$kill_rate,
     xlab = "Global kill rate",
     ylab = "test points",
     main = "Relationship between global Killrate and points")
abline(lm(points ~ kill_rate, data = tests), v = mean(tests$kill_rate))

```
# Submission timelines

TODO: Durch Anzahl Spieler teilen

```{r}
tmp <- tests |>
  mutate(seconds_since_gamestart = seconds_since_gamestart / 60) |>
  rename(`Minutes since gamestart` = seconds_since_gamestart)

ggplot(data = tmp, aes(x = `Minutes since gamestart`, colour = llm_or_human)) +
  geom_freqpoly(binwidth = 1)
```



```{r}
par(mfrow = c(2,2))
hist(
  tests[tests$llm_or_human == "Human", "seconds_since_gamestart"],
  breaks = 30,
  main = "human test submission timeline",
  xlab = "Seconds since game start"
)
hist(
  tests[tests$llm_or_human == "LLM", "seconds_since_gamestart"],
  breaks = 30,
  main = "llm test submission timeline",
  xlab = "Seconds since game start"
)
hist(
  tests[mutants$llm_or_human == "Human", "seconds_since_gamestart"],
  breaks = 30,
  main = "human mutant submission timeline",
  xlab = "Seconds since game start"
)
hist(
  tests[mutants$llm_or_human == "LLM", "seconds_since_gamestart"],
  breaks = 30,
  main = "llm mutant submission timeline",
  xlab = "Seconds since game start"
)
```
# llm messages until success

llm players can try to fix their generated tests/mutants if they failed validation.
These plots show the number of successfully submitted tests/mutants per minute,
grouped by the number of messages to the llm that were necessary for this submission.

Some tests have no directly associated number of messages. Those tests were generated
as part of the initial test suite and were successfully submitted on first try.


```{r}
tmp <- tests[tests$llm_or_human == "LLM",] |>
  mutate(number_of_messages = ifelse(is.na(number_of_messages), 0, number_of_messages)) |>
  mutate(number_of_messages = factor(number_of_messages,
                                     levels = c(0, 1,2,3,4),
                                     labels = c("Suite", "1", "2", "3", "4"))) |>
  mutate(seconds_since_gamestart = seconds_since_gamestart / 60) |>
  rename(`Minutes since gamestart` = seconds_since_gamestart) |>
  group_by(number_of_messages)


ggplot(data = tmp, aes(x = `Minutes since gamestart`, colour = number_of_messages)) +
  geom_freqpoly(binwidth = 1) + 
  ylab("tests per minute") +
  scale_colour_manual(name = "Number of messages", values = c("black", colors.quality))
  
```
```{r}
tmp <- mutants[mutants$llm_or_human == "LLM", ] |>
  mutate(number_of_messages = factor(number_of_messages,
                                     levels = c(0, 1,2,3,4),
                                     labels = c("Suite", "1", "2", "3", "4"))) |>
  mutate(seconds_since_gamestart = seconds_since_gamestart / 60) |>
  rename(`Minutes since gamestart` = seconds_since_gamestart) |>
  group_by(number_of_messages)


ggplot(data = tmp, aes(x = `Minutes since gamestart`, colour = number_of_messages)) +
  geom_freqpoly(binwidth = 1) + 
  ylab("mutants per minute") +
  scale_colour_manual(name = "Number of messages", values = c(colors.quality))
```



# Turing tests

llm-generated tests are almost never confused for human-written tests,
but mutants are hardly distinguishable.

```{r}
detach.dfs()
test_players <- tests[tests$opponent_llm_or_human == "Human" & !duplicated(tests[,c("game_id", "author_nickname")]) & tests$is_equivalence_test == "NO", c("game_id", "author_nickname", "llm_or_human", "opponent_opinion_human_or_ai")]
test_players <- test_players %>%
  mutate(opponent_opinion_human_or_ai = factor(opponent_opinion_human_or_ai, levels = 1:5, labels = interpretation.opponent_opinion_human_or_ai))

test_counts <- test_players %>% 
  count(llm_or_human, opponent_opinion_human_or_ai)

ggplot(data = test_counts, aes(x = llm_or_human, y = n, fill = opponent_opinion_human_or_ai)) + 
  geom_col(position = "fill", name = "Proportion") +
  scale_fill_manual(name = "Attacker's guess", values = colors.opinion) +
  xlab("Defender") +
  ylab("Proportion") +
  ggtitle("Defender Turing test")

```

```{r}
detach.dfs()
mutant_players <- mutants[mutants$opponent_llm_or_human == "Human" & !duplicated(mutants[,c("game_id", "author_nickname")]), c("game_id", "author_nickname", "llm_or_human", "opponent_opinion_human_or_ai")]
mutant_players <-mutant_players %>%
  mutate(opponent_opinion_human_or_ai = factor(opponent_opinion_human_or_ai, levels = 1:5, labels = interpretation.opponent_opinion_human_or_ai))
mutant_counts <-mutant_players %>% 
  count(llm_or_human, opponent_opinion_human_or_ai)

ggplot(data = mutant_counts, aes(x = llm_or_human, y = n, fill = opponent_opinion_human_or_ai)) + 
  geom_col(position = "fill", name = "Proportion") +
  scale_fill_manual(name = "Defender's guess", values = colors.opinion) +
  xlab("Attacker") +
  ylab("Proportion") + 
  ggtitle("Attacker Turing test")
rm(mutant_counts)
```
# stillborn mutants

A stillborn mutant is a mutant that was killed by a test that was already in the game.

## Proportions
This plot shows the proportions of stillborn mutants within a population.
We can see that llm-generated mutants are more likely to be stillborn than
human-written mutants, and that mutants are much more likely to be killed
on submission when playing against an llm player compared to a human player.

This last point can possibly be explained by the fact that llm defenders first 
generate a complete test suite before writing individual tests.

```{r}
stillborn_averages <- mutants %>%
  group_by(llm_or_human, opponent_llm_or_human) %>%
  summarise(
    stillborn = mean(stillborn)
  ) %>%
  mutate(
    stillborn = round(stillborn, digits = 3)
  )

ggplot(data=stillborn_averages, aes(x = llm_or_human, y = stillborn, fill = opponent_llm_or_human)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = percent(stillborn)), 
    position = position_dodge(0.9),
    vjust = 1.5,
  ) +
  scale_fill_manual(name = "Defender", values = colors.actor) + 
  xlab("Attacker") + 
  ylab("Percentage") + 
  ggtitle("Percentages of stillborn mutants") +
  scale_y_continuous(labels = percent)

rm(stillborn_averages)
```
There are, however, only very slight differences in total evasion rates
between stillborn and not-stillborn mutants.

```{r}
ggplot(data = mutants, aes(x = stillborn, y = evasion_rate, fill = llm_or_human)) +
  geom_boxplot(position = "dodge") +
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  scale_y_continuous(labels = percent, name = "Evasion rate") +
  scale_x_discrete(labels = c("Not stillborn", "stillborn"),) +
  xlab(NULL)
  ggtitle("Total evasion rates of stillborn and not-stillborn mutants")
```



# Equivalences
The following plots show the percentage of mutants that could not be killed by any test
in the entire study among their respective population.

We can see that at least 35%-40% of mutants that were accepted as equivalent
were actually killable by other tests in the study. Humans were only slightly
```{r}

tmp <- mutants |>
  group_by(
    llm_or_human,
    declared_equivalent,
    opponent_llm_or_human
  ) |>
  summarise(
    not_killed = mean(!has_been_killed)
  ) |>
  mutate(
    opponent_llm_or_human = paste(opponent_llm_or_human, "Defender")
  )

ggplot(
  data = tmp, 
  aes(x = declared_equivalent, y = not_killed, fill = llm_or_human)
  ) +
  geom_col(position = "dodge") +
  geom_text(aes(
    label = percent(not_killed),
    vjust = -0.4
    ), position = position_dodge(0.9)) +
  facet_wrap( ~ opponent_llm_or_human,) + 
  scale_y_continuous(labels = percent, name = "Mutants not killed by any test") + 
  scale_x_discrete(labels = c("Not declared equivalent", "Declared equivalent")) +
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  ggtitle("Equivalent mutants") + 
  xlab(NULL)

  
```
```{r}
tmp <- mutants |>
  group_by(
    llm_or_human,
    opponent_llm_or_human,
    has_been_killed
  ) |>
  summarise(
    declared_equivalent = mean(declared_equivalent)
  ) |>
  mutate(
    opponent_llm_or_human = paste(opponent_llm_or_human, "Defender")
  )

ggplot(
  data = tmp, 
  aes(x = has_been_killed, y = declared_equivalent, fill = llm_or_human)
  ) +
  geom_col(position = "dodge") +
  geom_text(aes(
    label = percent(declared_equivalent),
    vjust = -0.4
    ), position = position_dodge(0.9)) +
  facet_wrap( ~ opponent_llm_or_human,) + 
  scale_y_continuous(labels = percent, name = "Mutants declared as equivalent") + 
  scale_x_discrete(labels = c("Not killed by any test", "Killed at least once")) +
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  ggtitle("Equivalent mutants") + 
  xlab(NULL)
```

