---
title: "Stats for llm-Players"
output:
  html_notebook: default
  pdf_document: default
---
# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r}
library(ggplot2)
library(dplyr)
library(viridisLite)
library(tidyr)
library(scales)
getwd()
detach.dfs <- function() {
  for (p in search()) {
    if (grepl("tests|mutants", p)) {
      detach(p, character.only = TRUE)
    }
  }
}

colors.opinion <- c("red", "orange", "grey", "cyan", "blue")
colors.actor <- c("turquoise3", "thistle4")
colors.quality <- c("#458B00", "darkgoldenrod3", "#8B1C62", "red4")

interpretation.opponent_opinion_human_or_ai <- as.factor(c(
  "Definitely human",
  "Probably human",
  "Don't know",
  "Probably ai",
  "Definitely ai"
))

levels.degree <- c(
  "Bachelor Informatik", 
  "Bachelor Internet Computing", 
  "Lehramt Informatik",
  "Artificial Intelligence")
levels.cut <- c(
  "ByteVector",
  "CharRange"
)

common_variables <- function(.data) {
  mutate(
    .data,
    llm_or_human = factor(llm_or_human, levels = c("LLM", "Human")),
    opponent_llm_or_human = factor(opponent_llm_or_human, levels = c("LLM", "Human")),
    cut = factor(cut, levels = levels.cut),
    author_degree = factor(author_degree, levels = levels.degree),
    opponent_degree = factor(opponent_degree, levels = levels.degree),
    experiment_round = factor(experiment_round),
    author_opinion_judged_programming_skill = factor(author_opinion_judged_programming_skill),
    author_opinion_judged_original = factor(author_opinion_judged_original),
    author_opinion_challenged = factor(author_opinion_challenged),
    author_opinion_fun = factor(author_opinion_fun),
    author_opinion_human_or_ai = factor(author_opinion_human_or_ai),
    opponent_opinion_judged_programming_skill = factor(opponent_opinion_judged_programming_skill),
    opponent_opinion_judged_original = factor(opponent_opinion_judged_original),
    opponent_opinion_challenged = factor(opponent_opinion_challenged),
    opponent_opinion_fun = factor(opponent_opinion_fun),
    opponent_opinion_human_or_ai = factor(opponent_opinion_human_or_ai),
  )
}
```

```{r}
tests <- read.csv("test_data.csv") |>
  common_variables() |>
  mutate(
    is_equivalence_test = is_equivalence_test == "YES",
    kill_rate = (human_mutants_killed + llm_mutants_killed) / (human_mutants_killed + human_mutants_ignored + llm_mutants_killed + llm_mutants_ignored) |>
      replace_na(0),
    human_kill_rate = (human_mutants_killed / (human_mutants_killed + human_mutants_ignored)) |>
      replace_na(0),
    llm_kill_rate = (llm_mutants_killed / (llm_mutants_killed + llm_mutants_ignored)) |>
      replace_na(0),
    existing_kill_rate = (existing_mutants_killed / (existing_mutants_killed + existing_mutants_ignored)) |>
      replace_na(0),
    future_kill_rate = (future_mutants_killed / (future_mutants_killed + future_mutants_ignored)) |>
      replace_na(0),
  )

mutants <- read.csv("mutant_data.csv") |>
  common_variables() |>
  mutate(
    stillborn = stillborn == "YES",
    survived = survived == "YES",
    declared_equivalent = declared_equivalent == "YES",
    has_been_killed = human_tests_killed_by + llm_tests_killed_by > 0,
    
    evasion_rate = (human_tests_survived + llm_tests_survived) / (human_tests_survived + human_tests_killed_by + llm_tests_survived + llm_tests_killed_by) |>
      replace_na(0),
    human_evasion_rate = (human_tests_survived / (human_tests_survived + human_tests_killed_by)) |>
      replace_na(0),
    llm_evasion_rate = (llm_tests_survived / (llm_tests_survived + llm_tests_killed_by)) |>
      replace_na(0),
    existing_evasion_rate = (existing_tests_survived / (existing_tests_survived + existing_tests_killed_by)) |>
      replace_na(0),
    future_evasion_rate = (future_tests_survived / (future_tests_survived + future_tests_killed_by)) |>
      replace_na(0),
  )
  
number_of_test_games <- tests[!tests$is_equivalence_test,] |>
  group_by(
    llm_or_human,
  ) |>
  summarise(
    n = n_distinct(game_id)
  )
number_of_mutant_games <- mutants |>
  group_by(
    llm_or_human,
  ) |>
  summarise(
    n = n_distinct(game_id)
  )

deftests <- tests |> filter(!is_equivalence_test)
```

# Killrates

A "Killrate" is calculated for every single test. It describes the percentage of 
mutants within a specific population that would be killed by this test among 
those that are covered by this test.

"existing" and "future" mutants refers to those mutants within the same game
as the test that were submitted before/after the test was submitted.

```{r}

tmp <- tests |>
  pivot_longer(
    cols = c(kill_rate, llm_kill_rate, human_kill_rate, existing_kill_rate, future_kill_rate),
    names_to = "kill_rate_type",
    values_to = "kill_rate_value"
  ) |>
  mutate(kill_rate_type = factor(kill_rate_type, 
                                 levels = c(
                                   "kill_rate", 
                                   "human_kill_rate", 
                                   "llm_kill_rate", 
                                   "existing_kill_rate",
                                   "future_kill_rate"),
                                 labels = c(
                                   "Total kill rate",
                                   "Against human mutants",
                                   "Against llm mutants",
                                   "Against existing mutants",
                                   "Against future mutants"
                                 )))

ggplot(data = tmp, 
       aes(x = llm_or_human, y = kill_rate_value, fill = opponent_llm_or_human)
       ) + 
  geom_boxplot(position = "dodge") + 
  xlab("Defender") +
  scale_y_continuous(name = "Kill rate", labels = percent) + 
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  facet_wrap( ~ kill_rate_type)
```
# Relationship between points and killrates

While there is no clear pattern for the relationship between the points
earned in-game by test and their total "kill-rate" against all mutants submitted
in the study, mutants that earn many points tend to have higher "evasion-rates"
against all tests submitted in the study.

Mutants that were declared as equivalent while not being killed by any test
were excluded from the sample.

```{r}
tmp <- tests |> 
  filter(!is_equivalence_test,) |>
  mutate(
    point_ntil = ntile(points, n = 10)
  ) |>
  group_by(point_ntil)

ggplot(tmp, aes(y = kill_rate, x = factor(point_ntil), fill = llm_or_human)) +
  geom_boxplot() +
  scale_fill_manual(values = colors.actor, name = "Defender") + 
  ggtitle("Kill rates vs Points for tests") +
  xlab("Point deciles") + 
  ylab("Total kill rate") +
  scale_y_continuous(labels = percent)

```

```{r}
tmp <- mutants |> 
  filter(!(!has_been_killed &declared_equivalent)) |>
  mutate(
    point_ntil = ntile(points, n = 10)
  ) |>
  group_by(point_ntil)

ggplot(tmp, aes(y = evasion_rate, x = factor(point_ntil), fill = llm_or_human)) +
  geom_boxplot() +
  scale_fill_manual(values = colors.actor, name = "Attacker") + 
  ggtitle("Evasion rates vs Points for mutants") +
  xlab("Point deciles") + 
  ylab("Total evasion rate") +
  scale_y_continuous(labels = percent)
```

```{r}
ggplot(mutants, aes(x = points, y = evasion_rate)) +
  geom_point() + 
  geom_smooth(formula = y ~x, method = loess) +
  scale_y_continuous(limits = c(0,1), labels = percent, name = "Global evasion rate") + 
  xlab("Points")
```

# Submission timelines

These plots show the average number of tests/mutants submitted per minute per game.

We can see that LLM defenders submitted significantly more tests per minute than
human defenders, showing that the cooldown period was configured to be too low.

Submission rates for LLM defenders go down with time. It seems that the initial
"test suite" tests require re-generation less often than the uniquely written tests.

```{r}
timelineplot <- function(.data) {
  .data |>
  mutate(seconds_since_gamestart = floor(seconds_since_gamestart / 60)) |>
    rename(`Minutes since gamestart` = seconds_since_gamestart) |>
    mutate(
      part = as.numeric(ifelse(llm_or_human == "LLM",
                    1 / number_of_test_games[number_of_test_games$llm_or_human == "LLM","n"],
                    1 / number_of_test_games[number_of_test_games$llm_or_human == "Human","n"]
                    ))
    ) |>
    group_by(`Minutes since gamestart`, llm_or_human) |>
    summarise(
      part = sum(part)
    ) |>
  ggplot(aes(x = `Minutes since gamestart`, y = part, colour = llm_or_human)) +
    geom_line()
}

tests |>
  filter(!is_equivalence_test) |>
  timelineplot() +
  ylab("Test submissions per game and minute") +
  scale_colour_manual(values = colors.actor, name = "Defender")
```

Submission rates for mutants go down over time for both LLM and Human Attackers.
Since LLM attacker attempt to submit a new mutant every 30 seconds, this must
be related to higher rejection rates. A possible reason for that is the requirement
of mutants to be unique, there will be more collisions later in the game.

```{r}
mutants |>
  timelineplot() +
  ylab("Mutant submissions per game and minute") +
  scale_colour_manual(values = colors.actor, name = "Attacker")
```

# LLM messages until success

LLM players can try to fix their generated tests/mutants if they failed validation.
These plots show the number of successfully submitted tests/mutants per minute,
grouped by the number of messages to the LLM that were necessary for this submission.

Some tests have no directly associated number of messages. Those tests were generated
as part of the initial test suite and were successfully submitted on first try.


```{r}
tmp <- tests[tests$llm_or_human == "LLM",] |>
  mutate(number_of_messages = ifelse(is.na(number_of_messages), 0, number_of_messages)) |>
  mutate(number_of_messages = factor(number_of_messages,
                                     levels = c(0, 1,2,3,4),
                                     labels = c("Suite", "1", "2", "3", "4"))) |>
  mutate(seconds_since_gamestart = seconds_since_gamestart / 60) |>
  rename(`Minutes since gamestart` = seconds_since_gamestart) |>
  group_by(number_of_messages)


ggplot(data = tmp, aes(x = `Minutes since gamestart`, colour = number_of_messages)) +
  geom_freqpoly(binwidth = 1) + 
  ylab("Tests per minute") +
  scale_colour_manual(name = "Number of messages", values = c("black", colors.quality))
  
```

```{r}
tmp <- mutants[mutants$llm_or_human == "LLM", ] |>
  mutate(number_of_messages = factor(number_of_messages,
                                     levels = c(0, 1,2,3,4),
                                     labels = c("Suite", "1", "2", "3", "4"))) |>
  mutate(seconds_since_gamestart = seconds_since_gamestart / 60) |>
  rename(`Minutes since gamestart` = seconds_since_gamestart) |>
  group_by(number_of_messages)


ggplot(data = tmp, aes(x = `Minutes since gamestart`, colour = number_of_messages)) +
  geom_freqpoly(binwidth = 1) + 
  ylab("Mutants per minute") +
  scale_colour_manual(name = "Number of messages", values = c(colors.quality))
```



# Turing tests

LLM-generated tests are almost never confused for human-written tests,
but mutants are hardly distinguishable.

```{r}
tmp <- tests |>
  filter(opponent_llm_or_human == "Human", !is_equivalence_test) |>
  group_by(game_id, author_nickname, opponent_opinion_human_or_ai, llm_or_human) |>
  mutate(opponent_opinion_human_or_ai = factor(opponent_opinion_human_or_ai, levels = 1:5, labels = interpretation.opponent_opinion_human_or_ai)) |>
  summarise(n = n())

ggplot(data = tmp, aes(x = llm_or_human, y = n, fill = opponent_opinion_human_or_ai)) + 
  geom_col(position = "fill", name = "Proportion") +
  scale_fill_manual(name = "Attacker's guess", values = colors.opinion) +
  xlab("Defender") +
  ylab("Proportion") +
  ggtitle("Defender Turing test") + 
  scale_y_continuous(labels = percent)

```

```{r}
mutant_players <- mutants[mutants$opponent_llm_or_human == "Human" & !duplicated(mutants[,c("game_id", "author_nickname")]), c("game_id", "author_nickname", "llm_or_human", "opponent_opinion_human_or_ai")]
mutant_players <-mutant_players %>%
  mutate(opponent_opinion_human_or_ai = factor(opponent_opinion_human_or_ai, levels = 1:5, labels = interpretation.opponent_opinion_human_or_ai))
mutant_counts <-mutant_players %>% 
  count(llm_or_human, opponent_opinion_human_or_ai)

ggplot(data = mutant_counts, aes(x = llm_or_human, y = n, fill = opponent_opinion_human_or_ai)) + 
  geom_col(position = "fill", name = "Proportion") +
  scale_fill_manual(name = "Defender's guess", values = colors.opinion) +
  scale_y_continuous(labels = percent) +
  xlab("Attacker") +
  ylab("Proportion") + 
  ggtitle("Attacker Turing test")
rm(mutant_counts)
```
# Stillborn mutants

A stillborn mutant is a mutant that was killed by a test that was already in the game.

## Proportions
This plot shows the proportions of stillborn mutants within a population.
We can see that LLM-generated mutants are more likely to be stillborn than
human-written mutants, and that mutants are much more likely to be killed
on submission when playing against an LLM player compared to a human player.

This last point can possibly be explained by the fact that LLM defenders first 
generate a complete test suite before writing individual tests.

```{r}
tmp <- mutants %>%
  group_by(llm_or_human, opponent_llm_or_human) %>%
  summarise(
    stillborn = mean(stillborn)
  ) %>%
  mutate(
    stillborn = round(stillborn, digits = 3)
  )

ggplot(data=tmp, aes(x = llm_or_human, y = stillborn, fill = opponent_llm_or_human)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = percent(stillborn)), 
    position = position_dodge(0.9),
    vjust = 1.5,
  ) +
  scale_fill_manual(name = "Defender", values = colors.actor) + 
  xlab("Attacker") + 
  ylab("Percentage") + 
  ggtitle("Percentages of stillborn mutants") +
  scale_y_continuous(labels = percent)

```
There are, however, only very slight differences in total evasion rates
between stillborn and not-stillborn mutants.

```{r}
ggplot(data = mutants, aes(x = stillborn, y = evasion_rate, fill = llm_or_human)) +
  geom_boxplot(position = "dodge") +
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  scale_y_continuous(labels = percent, name = "Evasion rate") +
  scale_x_discrete(labels = c("Not stillborn", "Stillborn"),) +
  xlab(NULL) +
  ggtitle("Total evasion rates of stillborn and not-stillborn mutants")
```



# Equivalences
The following plots show the percentage of mutants that could not be killed by any test
in the entire study among their respective population.

For the purpose of this analysis, we assume that all mutants that could not be killed
by any test in the study are actually equivalent. 

We can see that at least 20% of human-written mutants that were declared as equivalent
after being claimed as equivalent by another human were actually killable. 

In contrast, at least 57% of LLM-written tests that 
were accepted as equivalent after being claimed as equivalent by a human were
killable. 

In LLM-vs-LLM-games, almost every equivalent test was marked as equivalent.
This is because in LLM-vs-LLM games, after the normal time ends, players
stop the normal test/mutant submission, the defender claims every surviving
mutant as equivalent, and the attacker is given the opportunity to defend
against that. This procedure was chosen because without it, many equivalent
tests with just shy of 10 points would in total provide a massive point 
advantage to the attacker. LLM-vs-LLM games are therefore not comparable to the
other combinations.
```{r}

tmp <- mutants |>
  group_by(
    llm_or_human,
    declared_equivalent,
    opponent_llm_or_human
  ) |>
  summarise(
    not_killed = mean(!has_been_killed)
  ) |>
  mutate(
    opponent_llm_or_human = paste(opponent_llm_or_human, "Defender")
  )

ggplot(
  data = tmp, 
  aes(x = declared_equivalent, y = not_killed, fill = llm_or_human)
  ) +
  geom_col(position = "dodge") +
  geom_text(aes(
    label = percent(not_killed),
    vjust = -0.4
    ), position = position_dodge(0.9)) +
  facet_wrap( ~ opponent_llm_or_human,) + 
  scale_y_continuous(labels = percent, name = "Equivalent mutants") + 
  scale_x_discrete(labels = c("Not declared\n equivalent", "Declared equivalent")) +
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  xlab(NULL)

  
```
```{r}
tmp <- mutants |>
  group_by(
    llm_or_human,
    opponent_llm_or_human,
    has_been_killed
  ) |>
  summarise(
    declared_equivalent = mean(declared_equivalent)
  ) |>
  mutate(
    opponent_llm_or_human = paste(opponent_llm_or_human, "Defender")
  )

ggplot(
  data = tmp, 
  aes(x = has_been_killed, y = declared_equivalent, fill = llm_or_human)
  ) +
  geom_col(position = "dodge") +
  geom_text(aes(
    label = percent(declared_equivalent),
    vjust = -0.4
    ), position = position_dodge(0.9)) +
  facet_wrap( ~ opponent_llm_or_human,) + 
  scale_y_continuous(labels = percent, name = "Mutants declared as equivalent") + 
  scale_x_discrete(labels = c("Equivalent", "Not equivalent")) +
  scale_fill_manual(name = "Attacker", values = colors.actor) +
  ggtitle("Equivalent mutants") + 
  xlab(NULL)
```

